On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-10-30 manuscript/pnas_revision.tex
ready to ship

In \cref{sec:sars:cov:app} we follow these theoretical investigations
with an application of the naive, first-order gradient approximation
\eqref{eq:firstOrder} to a challenge in phylogeography requiring
Bayesian inference of a rate-matrix of unprecedented dimensionality.
Namely, we apply the approximation to a gold-standard mixed-effects,
phylogenetic CTMC model that uses 1,897 parameters to describe the
spread of SARS-CoV-2 across a $d=44$ dimensional state space
consisting of different global geographic locations.  Such an
application complements the empirical studies of
\cite{magee2023random} in a manner that emphasizes the naive
approximation's potential for impact.

@@

We formulate our second major result as follows. 
\begin{theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d, \RR)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      $O_{\textnormal{a.s.}}$ is as in
      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{theorem}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}.  The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).  There appears to be no
  significant departure from the idea that the form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

@@

\section{Empirical Studies}

@@

Given $\phylogeny$, we model the evolution of $d$ characters along
each branch of the tree according to a CTMC model with $d\times d$
generator matrix $\QQ$ and stationary distribution
$\ppi=(\pi_1,\dots,\pi_d)=\lim_{t\rightarrow
  \infty}\widetilde{\ppi}e^{t\QQ}$, for $\widetilde{\ppi}$ for any
arbitrary probability vector.  Examples of characters are the $d=4$
nucleotides within a set of aligned genome sequences
\cite{jukes1969evolution} and the set of $d=15$ geographic regions
visited by an influenza subtype \cite{lemey2014unifying}.  We may
scale $t_v$ to be raw time (e.g., years) or the expected number of
substitutions with respect to $\ppi$ depending on the given problem.
In the former case, one may augment the model with a rate scalar
$\gamma$ that modulates the expected number of substitutions across
all branches, and the finite-time transition probability matrix along
branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.  In the following, we
further posit $\QQ=\QQ(\ttheta)$ for $\ttheta$ a vector of parameters.

@@

Lets briefly recall this HMC approach to resolve a given density
$\pi(\cdot)$ of a `target' probability measure of
interest.\footnote{In the Bayesian inference context, we are typically
  considering the posterior density function
  $\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
  $\ttheta\in\mathbb{R}^K$ given observed data $\Y$.}  We proceed by
considering a `potential energy' of the form
$U(\ttheta):=-\log \pi(\ttheta)$.  We then select an associated
potential energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
chosen symmetric-positive-definite mass matrix $\G$ (which is often
taken as the identity for simplicity).  One then observes that the
Gibbs measure, proportional to $e^{-H(\ttheta, \mom)}$ where
$H = U +V$, is invariant under the associated Hamiltonian dynamic.
Here note that the $\ttheta$ marginal of $e^{-H(\ttheta, \mom)}$
coincides with $\pi$ while the $\mom$ marginal is normally distributed
as $\mathcal{N}(\Zero,\G)$.

>>>>

 \cref{sec:emp}, contains simulation studies comparing: accuracy of matrix exponential derivative approximations for different distributional assumptions on the generator matrix; the posterior distributions obtained using surrogate-trajectory Hamiltonian Monte Carlo \cref{sec:Sur:trad:HMC} and traditional Hamiltonian Monte Carlo; and parameter identification under different priors on generator matrix elements.

In \cref{sec:sars:cov:app} we follow these theoretical and empirical investigations
with an application of the naive, first-order gradient approximation
\eqref{eq:firstOrder} to a challenge in phylogeography requiring
Bayesian inference of a rate-matrix of unprecedented dimensionality.
Namely, we apply the approximation to a gold-standard mixed-effects,
phylogenetic CTMC model that uses 1,897 parameters to describe the
spread of SARS-CoV-2 across a $d=44$ dimensional state space
consisting of different global geographic locations.  Such an
application complements the empirical studies of
\cite{magee2023random} in a manner that emphasizes the naive
approximation's potential for impact.

@@

We formulate our second major result as follows. 
\begin{theorem}\label{t:lambda_n-1(Q)->-infty_non-symm_a.s.}
  Let $F_X \in L_{\psi_1}$ be a distribution such that $X \geq 0$
  a.s., $\bbE X = \mu > 0$ and $\Var X = \sigma^2 > 0$. We consider a
  sequence of random matrices
  $\QQ \equiv \QQ(d) = \{q_{ij}\}_{i,j=1,\hdots,d} \in \mathcal{R}(d)$, $d \in \bbN$,
  where either
  \begin{align}\label{e:RM_theorem_general_condition}
    \begin{aligned}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i \neq j}
    \stackrel{\textnormal{iid}}\sim F_X
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d
  \end{aligned}
  \end{align}
  or we impose that $\QQ \in \mathcal{R}(d) \cap \mathcal{S}(d)$ as 
    \begin{align}
    \begin{aligned}\label{e:RM_theorem_symmetry_condition}
  &\{q_{ij}\}_{i , j =1, \hdots, d, \hspace{0.75mm}i > j}
  \stackrel{\textnormal{iid}}\sim F_X,
  q_{ij} := q_{ji} \text{ for } i < j
    \text{ and }\\
  &- q_{ii} = \sum_{j \in \{1,\hdots,d\} \backslash \{i\}}q_{ij},
  i=1,\hdots,d.
  \end{aligned}
\end{align}
    Then, in either of these cases, for any $d \in \bbN \backslash\{1\}$, we have
    \begin{align}
 \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big) \leq& \frac{\sigma_2(\QQ)}{d} 
 \notag \\
 \leq& \frac{\sigma_d(\QQ)}{d}
 \leq \mu+ O_{\as}\Big(\sqrt{ \frac{\log d}{d}}\Big)
 \label{e:lambda_n-1(Q)->-infty_non-symm}
      \end{align}
      almost surely. In \eqref{e:lambda_n-1(Q)->-infty_non-symm},
      $O_{\textnormal{a.s.}}$ is as in
      \eqref{e:X=O_a.s.(1)_Y=o_a.s.(1)}.
\end{theorem}

@@

\begin{remark}\label{rmk:non:sym:num:ev}
  Our rigorous formulation of \cref{cor:rig:det:rand:wrapup} is
  limited to symmetric random rate matrices whose above diagonal
  elements are independent and identically distributed draws from a
  sub-exponential distribution. However, strong numerical evidence
  suggest that the scope of the approximations
  \eqref{eq:bride:of:frankinstein:1:rm:app},
  \eqref{eq:bride:of:frankinstein:0:rm:app} reach far beyond the
  limitations of \cref{cor:rig:det:rand:wrapup} in several different
  ways. \cref{fig:app} explores the
  consequences of relaxing various assumptions of
  \cref{cor:rig:det:rand:wrapup}.  The third plot involves folded Cauchy random variables, the heavy tails of which violate the sub-exponential assumption (\cref{def:sub:exp:RV}).  There appears to be no
  significant departure from the idea that the form
  $t(\II -\QQ^+ \QQ) \JJ \QQ \QQ^{+}$ corresponds increasingly well to
  the true approximation error (\eqref{eq:true:diff}) as the dimension
  increases.
\end{remark}

@@

\section{Empirical Studies}\label{sec:emp}

@@

Given $\phylogeny$, we model the evolution of $d$ characters along
each branch of the tree according to a CTMC model with $d\times d$
generator matrix $\QQ$ and stationary distribution
$\ppi=(\pi_1,\dots,\pi_d)=\lim_{t\rightarrow
  \infty}\widetilde{\ppi}e^{t\QQ}$, for $\widetilde{\ppi}$ any
arbitrary probability vector.  Examples of characters are the $d=4$
nucleotides within a set of aligned genome sequences
\cite{jukes1969evolution} and the set of $d=15$ geographic regions
visited by an influenza subtype \cite{lemey2014unifying}.  We may
scale $t_v$ to be raw time (e.g., years) or the expected number of
substitutions with respect to $\ppi$ depending on the given problem.
In the former case, one may augment the model with a rate scalar
$\gamma$ that modulates the expected number of substitutions across
all branches, and the finite-time transition probability matrix along
branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.  In the following, we
further posit $\QQ=\QQ(\ttheta)$ for $\ttheta$ a vector of parameters.

@@

Lets briefly recall this HMC approach to resolve a given density
$\pi(\cdot)$ of a `target' probability measure of
interest.\footnote{In the Bayesian inference context, we are typically
  considering the posterior density function
  $\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
  $\ttheta\in\mathbb{R}^K$ given observed data $\Y$.}  We proceed by
considering a potential energy of the form
$U(\ttheta):=-\log \pi(\ttheta)$.  We then select an associated
kinetic energy $V(\mom) := | \G^{-1/2} \mom|^2$ for an appropriately
chosen symmetric-positive-definite mass matrix $\G$ (which is often
taken as the identity for simplicity).  One then observes that the
Gibbs measure, proportional to $e^{-H(\ttheta, \mom)}$ where
$H = U +V$, is invariant under the associated Hamiltonian dynamic.
Here note that the $\ttheta$ marginal of $e^{-H(\ttheta, \mom)}$
coincides with $\pi$ while the $\mom$ marginal is normally distributed
as $\mathcal{N}(\Zero,\G)$.
